---
knit: (function(input_file, encoding) {
  out_dir <- '.';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'Growth_rates_explanatory.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```
## 
```{r, echo=FALSE, include=FALSE}

#   ____________________________________________________________________________
#   Analysis                                                                ####


library(ape)
library(LineageHomology)
library(skygrowth)
library(grid)
library(scales) #Make more axis ticks.
library(gridExtra)
library(ggplot2)
library(phylodyn)

#Notes:
#Had to install an older version of RcppArmadillo to get the script running
#"install.packages("https://cran.r-project.org/src/contrib/Archive/RcppArmadillo/RcppArmadillo_0.9.900.3.0.tar.gz", repos=NULL, type="source")"

#Setup paths
path_main = "~/Dropbox/10000_Lovers/Phylogeography"
path_data = "~/Dropbox/10000_Lovers/01_LineageHomology/11_saved_data/"
path_results = "~/Dropbox/10000_Lovers/01_LineageHomology/12_saved_results/"

source("~/Dropbox/10000_Lovers/01_LineageHomology/sourcefunctions/environment_variables.R")

#RQ: Does meta-data explain the growth-rate of the biggest clades in period 2015-2018?
#1: Run Skygrowth on each TL.
#2: Extract clade wise growth-rate.
#3: Define explanatory variables for each clade: 
#4: Regress the growth rates on the metadata.


#   ____________________________________________________________________________
#   Set up treeclusters: Focus on the 20 largest from each location         ####
result = Result_NOR
clusters_and_times = function(result,main_tree,larger_than=10) {
  
  tree_clusters = list()
  time_shift_list = list() #For keeping track of time when shifting clusters. 
  for (i in 1:(which(result$Lineage_sizes<11)[1]-1))  #Loop over all lineages larger than 
  {
    cluster_tips = result$Taxa_names[i][[1]]
    tips_to_drop = main_tree$tip.label[main_tree$tip.label%in%cluster_tips==FALSE] #Tips to drop in this.
    mrca_node = getMRCA(main_tree,tip=cluster_tips)
    subtree = drop.tip(main_tree,tip =tips_to_drop)
    tree_clusters[[i]]= subtree
    time_shift_list[[i]] = nodeheight(main_tree, node=mrca_node)
  }
  return(list("clusters"=tree_clusters, "time_shifts"=time_shift_list))
}

neg_sizes = function(cluster_times) {
  neg_size_clusters = list()
  
  for(i in 1:length(cluster_times$time_shifts)) {
    tree_temp = cluster_times$clusters[[i]]
    class(tree_temp) = "phylo"
    neg_size_clusters[[i]]=skygrowth.map(tree_temp)
  }
  neg_size_clusters
}

custom_skygrowth_plot <- function( fit,shift_time=NA,ggplot=TRUE, logy=TRUE,xname=NA,xlimits=NA){
  ne <- fit$ne
  fit$time<-fit$time+shift_time
  pldf <- data.frame( t = fit$time,  nemed = ne, nelb = fit$ne_ci[,1], neub = fit$ne_ci[,3] )
  pl <- ggplot2::ggplot( pldf, ggplot2::aes_( x = ~ t, y = ~ nemed)) + ggplot2::geom_line()+ ggplot2::ylab('Effective population size') + ggplot2::xlab(xname)
  pl <- pl + ggplot2::geom_ribbon( ggplot2::aes_( ymin = ~ nelb, ymax = ~ neub), fill = 'blue', alpha = .2)
  if(logy==T){
    pl <- pl + ggplot2::scale_y_log10()  
  }
  if(is.na(xlimits)==F){
    ind_select = which(fit$time>=xlimits)
    fit$ne_ci[ind_select,]; ylims = range(fit$ne_ci[ind_select,])
    pl<-pl+lims(x=c(xlimits,NA),y=ylims)
  }
  return(pl)
}

skygrowth_multiplot = function(neg_sizes,root_date,time_shift_list,xlimits, logy) {
  c1 = lapply(1:length(neg_sizes), FUN = function(x) custom_skygrowth_plot(neg_sizes[[x]],shift_time=(root_date+time_shift_list[[x]]), xname=paste("Cluster: ",x),xlimits=xlimits, logy=logy))
  n <- length(c1)
  nCol <- floor(sqrt(n))
  do.call("grid.arrange", c(c1, ncol=nCol))
}

#Results for Norway
cluster_times_NOR = clusters_and_times(Result_NOR, main_tree = tree)
#neg_sizes_NOR = neg_sizes(cluster_times_NOR)
#save(neg_sizes_NOR,file=paste0(path_results,"neg_NOR.Rdata"))
load(file=paste0(path_results,"neg_NOR.Rdata"))
# pdf(paste0(path_results,"neg_NOR_top20.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_NOR,root_date=1750, time_shift_list=cluster_times_NOR$time_shifts,xlimits=2010, logy=F)
# dev.off()
# pdf(paste0(path_results,"neg_NOR_top20_full_timeline.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_NOR,root_date=1750, time_shift_list=cluster_times_NOR$time_shifts,xlimits=NA, logy=F)
# dev.off()

#Results for USA
cluster_times_USA = clusters_and_times(Result_USA, main_tree = tree)
#neg_sizes_USA = neg_sizes(cluster_times_USA)
#save(neg_sizes_USA,file=paste0(path_results,"neg_USA.Rdata"))
load(file=paste0(path_results,"neg_USA.Rdata"))
# pdf(paste0(path_results,"neg_USA_top20.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_USA,root_date=1750, time_shift_list=cluster_times_USA$time_shifts,xlimits=2010, logy=F)
# dev.off()
# pdf(paste0(path_results,"neg_USA_top20_full_timeline.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_USA,root_date=1750, time_shift_list=cluster_times_USA$time_shifts,xlimits=NA, logy=F)
# dev.off()

#Results for AUSTRALIA
cluster_times_AUS = clusters_and_times(Result_AUS, main_tree = tree)
#neg_sizes_AUS = neg_sizes(cluster_times_AUS)
#save(neg_sizes_AUS,file=paste0(path_results,"neg_AUS.Rdata"))
load(file=paste0(path_results,"neg_AUS.Rdata"))
# pdf(paste0(path_results,"neg_AUS_top20.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_AUS,root_date=1750, time_shift_list=cluster_times_AUS$time_shifts,xlimits=2010, logy=F)
# dev.off()
# pdf(paste0(path_results,"neg_AUS_top20_full_timeline.pdf"),width=15,height=15)
# skygrowth_multiplot(neg_sizes_AUS,root_date=1750, time_shift_list=cluster_times_AUS$time_shifts,xlimits=NA, logy=F)
# dev.off()

#Notes: Sampling is different between the locations
#The Australian collection, sampling stops in 2017 and starts in 2016 + some few genomes in 2004
#For USA there are a few samples from 1997, some from 2009, 2010, and the main bulk from 2016-2018, 2017 has much less than the neighboring years. 
#Thus we might extract the average growth rate in the period 2015-2017 for Australia, 2013-2018 for Norway, and  2009-2019 for USA? 


#   ___________________________________________________________________________________________
#   Can the growth rate be explained by explanatory variables such as sex distribution       ####


return_growth_interval=function(clusterwise_output, period=c(2010,2017), shift_time=1750,time_shift_list) {
  ncluster = length(clusterwise_output)
  average_growth_rate = c()
  for(i in 1:ncluster){
    times = clusterwise_output[[i]]$time+shift_time+time_shift_list$time_shifts[[i]] #Adding Root date + cluster tmrca.
    inds = which(times>=period[1]&times<=period[2]) #indexes in time periods.
    growth_rate = mean(clusterwise_output[[i]]$growthrate[inds])
    average_growth_rate=c(average_growth_rate,growth_rate)
  } 
  return(average_growth_rate)
}

#Extract the growth rates on the intervals 2014-2017: This may not be meaningful. 
r_NOR = return_growth_interval(clusterwise_output=neg_sizes_NOR, time_shift_list=cluster_times_NOR)
r_AUS = return_growth_interval(clusterwise_output=neg_sizes_AUS, time_shift_list=cluster_times_AUS)
r_USA = return_growth_interval(clusterwise_output=neg_sizes_USA, time_shift_list=cluster_times_USA)
r_all = c(r_NOR,r_AUS,r_USA)

define_explanatory_variables = function(result, larger_than) {
  #Sex distribution
  sex = sex_distributions(RLH = result,larger_than=larger_than)
  sex_variable = c()
  for(i in seq(1,nrow(sex),3)){ #Removes no data column.
    fem = sex$Count[i]
    mal = sex$Count[i+1]
    tmp=(mal)/sum(fem+mal)
    if(sum(fem+mal)<floor(larger_than/2)) { #If there are less than half of the minimum size we have nearly no information -> set Sex to NA. 
      tmp = NA  
    }
    sex_variable = c(sex_variable,tmp)
    
  }
  #MTR mosaic category: sets up a categorical variable.
  mtrDs = mtrD(RLH = result,larger_than=larger_than)
  mtr_variable = c()
  lineage_size = c() #defining this here for convenience.
  for(i in seq(1,nrow(mtrDs),7)){
    mtrDcounts = mtrDs$Count[i:(i+5)]
    maxind=which.max(mtrDcounts)
    mtr_variable=c(mtr_variable,mtrDs$State[i+(maxind-1)])
    lineage_size=c(lineage_size,mtrDs$Lineage_size[i+(maxind-1)]) #No reason to make it this complex (copypasted). 
  }
  
  #PenA state category
  penAs = penA_mosaic_distribution(RLH = result,larger_than=larger_than)
  penA_variable=c()
  for(i in seq(1,nrow(penAs),4)){
    penAscounts = penAs$Count[i:(i+3)]
    maxind=which.max(penAscounts)
    penA_variable=c(penA_variable,penAs$State[i+(maxind-1)])
  }
  d1 = data.frame(lineage_size,penA_variable,mtr_variable,sex_variable)
  return(d1)
}

d1 = define_explanatory_variables(Result_NOR,larger_than=10)
d2 = define_explanatory_variables(Result_AUS,larger_than=10)
d3 = define_explanatory_variables(Result_USA,larger_than=10)


#Gather data
total_data = rbind(d1,d2,d3)
locations = c(rep("Norway",length(r_NOR)),rep("Australia",length(r_AUS)),rep("USA",length(r_USA)))
total_data = cbind(r_all,locations,total_data) #Add growth rate
colnames(total_data)= c("GrowthRate","locations", "tlSize","penA","mtr", "sexDistribution") 

#   ____________________________________________________________________________
#   Report starts from here                                                 ####


```

## Analysis of the growth rate of the largest Tranmission lineages and their relation to explanatory variables.

##### Research question: Is the growth and size of transmission lineages in the time period 2010-2017 dependent on any of the explanatory variables: Sex distribution in the clade, mtD status, penA status, or final lineage size (transmissin lineage size)?.

The input for the following analyses are all the transmission lineages containing more than 10 sequences from each location: (Norway, Australia, USA). For each of the transmission lineages we run [skygrowth v.0.3.1](https://github.com/mrc-ide/skygrowth) on the clades, and extract the growth rate of the effective population size in the time interval 2010-2017

### Transmission lineages in Norway

```{r, echo=FALSE}

noquote("Size of transmission lineages containing more than 10 sequences:")
Result_NOR$Lineage_sizes[Result_NOR$Lineage_sizes>10]
noquote("Number of transmission containing more than 10 sequences:")
length(Result_NOR$Halfedge_over_tmrca[Result_NOR$Lineage_sizes>10])

``` 


With estimated arrival times (midpoint of the ancestral branch of the transmission lineage) in Norway:  

```{r, echo=FALSE}

noquote("Estimated arrival time in Norway")
Result_NOR$Halfedge_over_tmrca[Result_NOR$Lineage_sizes>10]

``` 

From skygrowth we find that the average growth rate of these effective population size of these lineage on the time interval 2010-2017 is: 
```{r, echo=FALSE}
noquote("Average growth rate in 2010-2017:")
r_NOR
``` 

The fourth to the last produces the value "NA", because this transmission lineage does not exist on the time interval [2010, 2017].

Similarly, for Australia and USA we find:

### Transmission lineages in Australia

```{r, echo=FALSE}
noquote("Size of transmission lineages containing more than 10 sequences:")
Result_AUS$Lineage_sizes[Result_AUS$Lineage_sizes>10]
noquote("Number of transmission containing more than 10 sequences:")
length(Result_AUS$Halfedge_over_tmrca[Result_AUS$Lineage_sizes>10])

noquote("Estimated arrival time in Australia")
Result_AUS$Halfedge_over_tmrca[Result_AUS$Lineage_sizes>10]

noquote("Average growth rate in 2010-2017:")
r_NOR
```

### Transmission lineages in USA

```{r, echo=FALSE}
noquote("Size of transmission lineages containing more than 10 sequences:")
Result_USA$Lineage_sizes[Result_USA$Lineage_sizes>10]
noquote("Number of transmission containing more than 10 sequences:")
length(Result_USA$Halfedge_over_tmrca[Result_USA$Lineage_sizes>10])

noquote("Estimated arrival time in USA")
Result_USA$Halfedge_over_tmrca[Result_USA$Lineage_sizes>10]

noquote("Average growth rate in 2010-2017:")
r_NOR

``` 

## Metadata

To study the relation of the growth rates to metadata we collect the growth rates in a data frame along with variables that describe the sex distribution, penA status and mtrD status for each transmission lineage.

The dataset now looks like
```{r, echo=FALSE}
head(total_data)
str(total_data)

```


## Modeling the growth rates and lineage sizes on the metadata


```{r, echo=FALSE, include = T}

library(grid)
g2 = ggplot(total_data)+theme_bw()+geom_point(aes(x=tlSize, y=GrowthRate,color=locations))+theme(legend.position = c(0.8,0.8))+ylab("Growth rate")+xlab("Transmission lineage size")
g2

```

Here we investigate the growth rate and size of the transmission lineages, and the relation of these to the explanatory variables: penA, mtrD, sex distribution and location. 

```{r, echo=FALSE, include = T, warning=F, message=F}
g1 = ggplot(total_data, aes(x=sexDistribution, y=GrowthRate, color=locations, fill=locations))+theme_bw()+geom_point()+ylab("Growth rate")+theme(legend.position = "none")+geom_smooth(method="lm",alpha=0.15)
g2 = ggplot(total_data)+theme_bw()+geom_boxplot(aes(x=locations, y=GrowthRate,color=locations))+theme(legend.position = "none")
g3 = ggplot(total_data)+theme_bw()+geom_boxplot(aes(x=mtr, y=GrowthRate))+ylab("Growth rate")
g4 = ggplot(total_data)+theme_bw()+geom_boxplot(aes(x=penA, y=GrowthRate))+ylab("Growth rate")
grid.arrange(g1,g2,g3,g4,nrow=2,ncol=2)
```

Lineage size versus variables explanatory variables

```{r, echo=FALSE, include = T, warning=F, message=F}
g1 = ggplot(total_data,aes(x=sexDistribution, y=tlSize, color=locations, fill=locations))+theme_bw()+geom_point()+ylab("Transmission lineage size")+theme(legend.position="none")+geom_smooth(method="lm", alpha=0.15)
g2 = ggplot(total_data)+theme_bw()+geom_boxplot(aes(x=locations, y=tlSize,color=locations))+theme(legend.position = "none")+ylab("Transmission lineage size")
g3 = ggplot(total_data)+theme_bw()+geom_boxplot(aes(x=mtr, y=tlSize))+ylab("Transmission lineage size")
g4 = ggplot(total_data)+theme_bw()+geom_boxplot(aes(x=penA, y=tlSize))+ylab("Transmission lineage size")
grid.arrange(g1,g2,g3,g4,nrow=2,ncol=2)

#boxplot(total_data$GrowthRate~as.factor(total_data$locations),ylab="Growth Rate", xlab="Transmission lineage size")
```

```{r, echo=FALSE, include = F}
total_data_modified = total_data #Modfied version of data
total_data_modified$penA[which(total_data_modified$penA=="nd")]=NA #set nd to NA
total_data_modified$penA[which(total_data_modified$penA=="semi-mosaic")]="mosaic" #set nd to NA
total_data_modified$mtr[which(total_data_modified$mtr=="nd")]=NA
total_data_modified$mtr[total_data_modified$mtr!="non-mosaic"]="mosaic"
total_data_modified$mtr[total_data_modified$mtr=="semi-mosaic"]="mosaic"
```

Linear regression of the lineage size and growth rate on the explanatory variables
*Note.The model fit is poor, should probably use GLM instead.*

```{r, echo=FALSE, include = T}
#Testing regression models
mod_lineage_size = lm(tlSize~penA+mtr+sexDistribution+locations, data = total_data)
summary(mod_lineage_size)

mod_growthRate = lm(GrowthRate~penA+mtr+sexDistribution+locations, data = total_data)
summary(mod_growthRate)

```

Collapsing the categories into two levels: Mosaic and non mosaic we have the data

```{r, echo=FALSE, include = T}
mod_lineage_size = lm(tlSize~penA+mtr+sexDistribution+locations, data = total_data_modified)
summary(mod_lineage_size)

mod_growthRate = lm(GrowthRate~penA+mtr+sexDistribution+locations, data = total_data_modified)
summary(mod_growthRate)

```


```{r, echo=FALSE, include = F}
mod_total = lm(r_all~factor(penA)+factor(mtr)+sexDistribution, data = total_data); summary(mod_total)

mod_total = lm(r_all~factor(penA)+factor(mtr)+sexDistribution, data = total_data_modified); summary(mod_total)

#plot(mod_total)

#plot(mod_total)
mod1 = lm(lineage_size~factor(penA_variable)+factor(mtr_variable)+sex_variable, data = d1)
mod2 = lm(lineage_size~factor(penA_variable)+factor(mtr_variable)+sex_variable, data = d2)
mod3 = lm(lineage_size~factor(penA_variable)+factor(mtr_variable)+sex_variable, data = d3)

#plot(mod1)

summary(mod1);
summary(mod2);
summary(mod3)


#Use functions from the app script to extract explanatory variables. 
# mtrD(Result,larger_than = input$largerthan)
# t1 = sex_distributions(Result,larger_than = input$largerthan)
# penA_mosaic_distribution(Result,larger_than = input$largerthan)

```



